Discrete Ordinates

### 2017-07-05 docprep/appdx_a.tex
Fixed typos and updated LaTeX template

The flux moments are an alternate representation of the angular flux  within a system. The rationale behind usage of flux moments is that they map the discrete flux ($\psi$) from a function of angle to a function of $P_N$ expansion. This reduces the amount of memory required for computation. Storing $\psi$ directly requires storing $G \times N_V \times N_a$ values where $G$ is the number of groups, $N_V$ is the number of voxels, and $N_a$ is the number of angles. Storing the flux moments, however, requires storing only $G \times N_V \times (N^2 + 2N + 1)$ values where $N$ is the $P_N$ expansion value. For example, a problem requiring $S_6$ needs $6 \times 8 = 48$ directions, but a problem requiring $P_6$ needs 49 expansion coefficients which would seem approximately eqivalent, but typical problems require far more discrete directions than expansion coefficients. Comples problems can easily require $S_{16}$ or higher but expansions above $P_5$ are rarely encountered.

@@

The Legendre polynomials are solutions to:
\begin{equation}
\begin{split}
P_0(\mu) &= 1 \\
P_l(\mu) &= \frac{1}{2^l l!} \frac{d^l}{d\mu^l}(\mu^2 - 1)^l \,, \quad l \in \mathbb{N}
\end{split}
\end{equation}
where $\mathbb{N}$ is the set of natural numbers (1, 2, 3...). Table~\ref{tab:legendre} gives the solution of the first few Legendre polynomials. The Legendre polynomials are orthogonal on the domain $[-1, 1]$ in that
\begin{equation}
\frac{1}{2} \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = 
\begin{cases}
\frac{1}{2l+1} \,, \quad l = l'\\
0 \,, \quad \text{otherwise}
\end{cases}
\end{equation}

@@

The spherical harmonics are defined as:
\begin{equation}
Y_{lm}(\hat{\Omega}) = \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P_l^m(\mu)e^{e \tau m}
\end{equation}
where $\mu$ is the cosine of the angle between $\hat{Omega}$ and the $x$-axis and $\tau$ is the rotation about the $x$-axis with respect to the $y$-axis of $\hat{\Omega}$ projected onto the $yz$ plane which can be compuated as:
\begin{equation}
\tau = \frac{\eta}{\sqrt{\eta^2 + \xi^2}}.
\end{equation}
The spherical harmonics have the orghogonality
\begin{equation}
\int_{}^{}Y_{lm}(\hat{Omega})Y_{l'm'}^*(\hat{\Omega}) d\hat{\Omega}= 
\begin{cases}
1 \,, \quad l = l' \text{ and } m = m' \\
0 \,, \quad \text{otherwise}
\end{cases}
\end{equation}
where $Y_{lm}^*$ is the complex conjugate of $Y_{lm}$. Using the addition theorem which states
\begin{equation}
P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2l+1}\sum_{m=-l}^{l}Y_{lm}^*(\hat{\Omega}') Y_{lm}(\hat{\Omega})
\end{equation}
gives
\begin{equation}
P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu') + 2\sum_{m=1}^{l}\frac{(l-m)!}{(l+m)!}P_l^m(\mu) P_l^m(\mu') cos(m[\tau - \tau']).
\end{equation}

>>>>

The flux moments are an alternate representation of the angular flux  within a system. The rationale behind usage of flux moments is that they map the discrete flux ($\psi$) from a function of angle to a function of $P_N$ expansion. This reduces the amount of memory required for computation. Storing $\psi$ directly requires storing $G \times N_V \times N_a$ values where $G$ is the number of groups, $N_V$ is the number of voxels, and $N_a$ is the number of angles. Storing the flux moments, however, requires storing only $G \times N_V \times (N^2 + 2N + 1)$ values where $N$ is the $P_N$ expansion value. For example, a problem requiring $S_6$ needs $6 \times 8 = 48$ directions, but a problem requiring $P_6$ needs 49 expansion coefficients which would seem approximately eqivalent, but typical problems require far more discrete directions than expansion coefficients. Complex problems can easily require $S_{16}$ or higher but expansions above $P_5$ are rarely encountered.

@@

The Legendre polynomials are solutions to:
\begin{equation}
\begin{split}
P_0(\mu) &= 1 \\
P_l(\mu) &= \frac{1}{2^l l!} \frac{d^l}{d\mu^l}(\mu^2 - 1)^l \,, \quad l \in \mathbb{N}
\end{split}
\end{equation}
where $\mathbb{N}$ is the set of natural numbers (1, 2, 3...). Table~\ref{tab:legendre} gives the solution of the first few Legendre polynomials. The Legendre polynomials are orthogonal on the domain $[-1, 1]$ in that
\begin{equation}
\frac{1}{2} \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = 
\begin{cases}
\frac{1}{2l+1} \,, \quad l = l'\\
0 \,, \quad \text{otherwise}
\end{cases}
\end{equation}
holds.

@@

The spherical harmonics are defined as:
\begin{equation}
Y_{lm}(\hat{\Omega}) = \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P_l^m(\mu)e^{e \tau m}
\end{equation}
where $\mu$ is the cosine of the angle between $\hat{\Omega}$ and the $x$-axis and $\tau$ is the rotation about the $x$-axis with respect to the $y$-axis of $\hat{\Omega}$ projected onto the $yz$ plane which can be compuated as:
\begin{equation}
\tau = \frac{\eta}{\sqrt{\eta^2 + \xi^2}}.
\end{equation}
The spherical harmonics have the orghogonality
\begin{equation}
\int_{}^{}Y_{lm}(\hat{\Omega})Y_{l'm'}^*(\hat{\Omega}) d\hat{\Omega}= 
\begin{cases}
1 \,, \quad l = l' \text{ and } m = m' \\
0 \,, \quad \text{otherwise}
\end{cases}
\end{equation}
where $Y_{lm}^*$ is the complex conjugate of $Y_{lm}$. Using the addition theorem which states
\begin{equation}
P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2l+1}\sum_{m=-l}^{l}Y_{lm}^*(\hat{\Omega}') Y_{lm}(\hat{\Omega})
\end{equation}
gives
\begin{equation}
P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu') + 2\sum_{m=1}^{l}\frac{(l-m)!}{(l+m)!}P_l^m(\mu) P_l^m(\mu') cos(m[\tau - \tau']).
\end{equation}

### 2017-07-05 docprep/chapbib.tex
Fixed typos and updated LaTeX template

\section{CT Dose Estimation}
As discussed in Chapter 1, the dose a patient receives from both diagnostic CT procedures and radiation therapy is of concern to the medical community. However, dose quantification is not currently patient specific. Measuring the dose in a patient directly is impossible, intead the air kerma (measured in Gy), which measures the x-ray beam intensity, and can be measured directly with ionization chambers~\citep{ef:wolbarsta}. As the x-ray beam passes through the patient, it deposits energy. The energy deposition per unit mass is the absorbed dose (measured in Gy) which is then weighted by the radiation type to give the equivalent dose (measured in Sv). The equivalent dose quantifies the biological detriment to the patient from the procedure. The effective dose (measured in Sv) accounts for the radiosensitivity of each organ and is the most meaningful measure of risk to the patient. Any two procedures that result in the same effective dose to the patient carry the same acute and long-term risks. Any procedure that gives an effective dose to a patient has the same risk as any procedure that would give the same whole body dose.

Unfortunately, neither the absorbed dose, equivalent dose, nor effective dose to a patient can be measured directly. Instead, the CT Dose Index (CTDI) is measured using a pencil shaped ionization chamber at the center of  a standardized acrylic phantom placed at the scanner isocenter~\citep{ref:hudaw}. The CTDI quantifies the equivalent dose the patient receives from a single slice of the CT scan protocol. The CTDI does \textit{not} capture patient-specific papameters, but rather quantifies the amount or radiation emitted by the source.

The CTDI does not provide the dose from an entire procedure, but rather a single slice. The dose from a procedure is obtained from the dose-length-product (DLP) which is the product of the CTDI and the axial scan length. The DLP is related to the effective dose to the patient for a particular procedure by a conversion factor that is dependent on many variables including the anatomical region scanned, the x-ray tube voltage, and the patient's age, gender, and size~\citep{ref:hudaw}.

INSERT MORE EXAMPLES.
\citet{ref:lauk} found that severly obese patients received twice as much dose from CT procedures as others on average. Some patients received as much as five times the typical dose.

@@

\citet{ref:ottossonr} extended the methodology provided by \citet{ref:schneideru} to provide 19 dosimetric groups of materials in the diagnostic energy domain independent of the particular scanner used to generate the data. The DOCTORS code relies on his proposed 19-group model. Table~\ref{table:ctmap} shows the mapping of CT number to material composition (reproduced from \citep{ref:ottossonr}).

@@

INSERT KBA DISCUSSION

@@

Many raytracing algorithms are based on the Bresenham line raserizing algorithm which was designed for graph plotting hardware~citep{ref:bresenhamj}. That algorithm was intended to produce lines acceptable for human interpretation and no guarantee is made that all pixels along the line will be identified making this class of algorithms inappropriate for dosimetric raytracing since attenuation through some voxels would be skipped. Some extensions have been added to it that guarantee passage through all voxels though~\citep{ref:liuy}.

@@

INSERT SECTION ON GPU RAYTRACING

@@

A graphics processing unit (GPU) is similar to a central processing unit (CPU)in that it executes instructions to process data. However, it differs in that a CPU may have up up to 22 physical cores, each of which can operate independently, a GPU has up to thousands of cores, but they cannot operate independently. Instead, cores execute simulaneously in groups called warps. All cores in a warp execute the same instruction at the same time as all others.

Two major GPU manufacturers control the vast majority of the market, Nvidia and AMD [CITE]. In 2007, Nvidia released its Compute Unified Device Architecture (CUDA) language. CUDA allows code written in C/C++/FORTRAN to communicate with a GPU with minimal additional code. Therefore, CUDA was selected for the GPU acceleration of DOCTORS. Therefore, only Nvidia GPUs are considered in this work.

@@

Monte Carlo dose computation for protons has been accelerated by GPU technology \citep{ref:jiax} and Monte Carlo in voxelized geometries has been done by \citep{ref:hissoinys} for monoenergetic beams.

>>>>

\section{CT Dose Estimation}
As discussed in Chapter 1, the dose a patient receives from both diagnostic CT procedures and radiation therapy is of concern to the medical community. However, dose quantification is not currently patient specific. Measuring the dose in a patient directly is impossible, instead the air kerma (measured in Gy), which measures the x-ray beam intensity, and can be measured directly with ionization chambers~\citep{ref:wolbarsta}. As the x-ray beam passes through the patient, it deposits energy. The energy deposition per unit mass is the absorbed dose (measured in Gy) which is then weighted by the radiation type to give the equivalent dose (measured in Sv). The equivalent dose quantifies the biological detriment to the patient from the procedure. The effective dose (measured in Sv) accounts for the radiosensitivity of each organ and is the most meaningful measure of risk to the patient. Any two procedures that result in the same effective dose to the patient carry the same acute and long-term risks. Any procedure that gives an effective dose to a patient has the same risk as any procedure that would give the same whole body dose.

Unfortunately, neither the absorbed dose, equivalent dose, nor effective dose to a patient can be measured directly. Instead, the CT Dose Index (CTDI) is measured using a pencil shaped ionization chamber at the center of  a standardized acrylic phantom placed at the scanner isocenter~\citep{ref:hudaw}. The CTDI quantifies the equivalent dose the patient receives from a single slice of the CT scan protocol. The CTDI does \textit{not} capture patient-specific papameters, but rather quantifies the amount of radiation emitted by the source.

The CTDI does not provide the dose from an entire procedure, but rather a single slice. The dose from a procedure is obtained from the dose-length-product (DLP) which is the product of the CTDI and the axial scan length. The DLP is related to the effective dose to the patient for a particular procedure by a conversion factor that is dependent on many variables including the anatomical region scanned, the x-ray tube voltage, and the patient's age, gender, and size~\citep{ref:hudaw}. For example, \citet{ref:lauk} found that severly obese patients received twice as much dose from CT procedures as others on average. Some patients received as much as five times the typical dose.

@@

\citet{ref:ottossonr} extended the methodology provided by \citet{ref:schneideru} to provide 19 dosimetric groups of materials in the diagnostic energy domain independent of the particular scanner used to generate the data. The DOCTORS code relies on his proposed 19-group model. Table~\ref{table:ctmap} shows the mapping of CT number to material composition (reproduced from \citet{ref:ottossonr}).

@@

@@

Many raytracing algorithms are based on the Bresenham line raserizing algorithm which was designed for graph plotting hardware~\citep{ref:bresenhamj}. That algorithm was intended to produce lines acceptable for human interpretation and no guarantee is made that all pixels along the line will be identified making this class of algorithms inappropriate for dosimetric raytracing since attenuation through some voxels would be skipped. Some extensions have been added to it that guarantee passage through all voxels though~\citep{ref:liuy}.

@@

@@

A graphics processing unit (GPU) is similar to a central processing unit (CPU)in that it executes instructions to process data. However, it differs in that a CPU may have up to 22 physical cores, each of which can operate independently, a GPU has up to thousands of cores, but they cannot operate independently. Instead, cores execute simulaneously in groups called warps. All cores in a warp execute the same instruction at the same time as all others.

Two major GPU manufacturers control the vast majority of the market, Nvidia and AMD. In 2007, Nvidia released its Compute Unified Device Architecture (CUDA) language. CUDA allows code written in C/C++/FORTRAN to communicate with a GPU with minimal additional code. Therefore, CUDA was selected for the GPU acceleration of DOCTORS. Therefore, only Nvidia GPUs are considered in this work.

@@

%Monte Carlo dose computation for protons has been accelerated by GPU technology \citep{ref:jiax} and Monte Carlo in voxelized geometries has been done by \citep{ref:hissoinys} for monoenergetic beams.

### 2017-07-05 docprep/chapconc.tex
Added some TODO
Fixed typos and updated LaTeX template

The collided flux suffers from the isotropic approximation made to simplify the code.

\section{Applicability to Clinical Settings}
With further development, DOCTORS is well positioned to become a code of significant clincal impact.

Great effort was put into producing a code that is simple to use yet capable of producing powerful results quickly. However, usage in a clinical setting would likely require further refinement of the GUI to make the code more robust and intuitive. Additionally, the direct output of DOCTORS may not necessarily be of clinical importance since

@@

The GPU acceleration algorithm was found to speed the DOCTORS code up by a factor of ~40x.

@@

A number of simple modifications to DOCTORS could be made that would greatly increase the code's usability and robustness. One of these modifications, generation of more refined group structures, is not a change to DOCTORS \textit{per se} but rather a change in the input cross section data. The other changes are additions to DOCTORS that would add new capabilities to the software.

In addition to specific modifications, some additional, broader future goals can also be identified.

\subsection{Anisotropy Treatment}
A more sophisticated anisotropy treatment would be appropriate since the collided flux was found to have issues regarding its behavior.

\subsection{Group Structure}

The cross section data currently used by DOCTORS are taken from SCALE6.2. While these cross sections have been found to be sufficient to produce flux distributions in medical CT imaging, a more refined group structure designed for medical applications would be worth investigation.

This work can be done using either NJOY [CITE] or the newly released AMPX [CITE] code. Either code has the capability to collapse an ENDF formatted data file into a group structure of the user's choosing in the format DOCTORS reads.

INSERT SECTION ABOUT NJOY or AMPX.

\subsection{Temporal Extension}
An advantage of the raytracer is that it is very fast. Therefore, it can characterize many beams and integrate them temporally easily. This would allow DOCTORS to model more complex beam shapes and scan protocols accurately.

\subsection{Partial Acceptance Criteria}
In the current version of doctors, the raytracer is very accurate with respect to MCNP6 in predicting the uncollided flux, except for along the periphery of the beam. Voxels are currently either completely inside the beam or completely outside of it as determined by its isocenter. A more sophisticated technique whereby a voxel cah be partially accepted in the beam would remove these artifacts.

\subsection{Multileaf Collimation}
Some medical systems currently utilize a multileaf collimator [CITE] to deliver dose to a specific organ or region of the body. As the

\subsection{Extension to Therapy Beams}
DOCTORS can likely be extended to high energy therapeutic beams for clinical treatment. However, higher energy photons will scatter more anisotropicly. This will require spherical harmonics and high order $P_l$ expansion tables in the cross section data.

\subsection{Organ Identification}
Currently, DOCTORS cannot automatically identify specific organs, thus it is only able to compute the equivalent dose via dose deposition. If specific organs could be identified, tissue specific weighting factors could be applied resulting in the effective dose to each organ which would be of greater clinical significance. However, such identification currently requires additional input about the scan protocol and extensive knowledge about the human anatomy not currently integrated into DOCTORS.

>>>>

Qualitatively, the most of the trends identified by DOCTORS agreed very well with MCNP6. The largest discrepancy was in the attenuation of particles through the patient region. However, since the raytracer is very accurate with respect to MCNP6, the difference is likely with the discrete ordinate methodology rather that cross section related. The key driver of this discrepancy is believed to be due to poor treatment of the angular transport either by way of the quadrature or anisotropy treatment.

This indicates that DOCTORS is better suited for problems that use a broad, distributed source as opposed to a narrow, directionally biased source. Such directional sources, require much higher quadratures to accurately characterize the beam due to ray effect.

\section{Applicability to Clinical Settings}
With further development, DOCTORS is well positioned to become a code of significant clincal impact. Though it is still lacking in medical diagnostic quality accuracy, it is easily extensible and capable of characterization of complex beams making it ideal for some types of clinical dosimetry.

Great effort was put into producing a code that is simple to use yet capable of producing powerful results quickly. However, usage in a clinical setting would likely require further refinement of the GUI to make the code more robust and intuitive. Additionally, the direct output of DOCTORS may not necessarily be of clinical importance since radiologists and technitians are more interested in compliance with regulations and ensuring patient safety. Therefore, a bulk report of overall numbers would likely be of more benfit than the entire spatial flux and dose distribution in a clinical setting.

@@

The GPU algorithm was implemented in CUDA and is very straightforward. At each subsweep, the global index of all voxels that can be computed in parallel are determined and a kernel is launched that executes those voxels. Unsurprisingly, this algorithms scales much better on larger problems where more time is spent in the computational execution of the kernal as opposed to the overhead operations of launching the kernel and memory transfer.

The GPU acceleration algorithm was found to speed the DOCTORS code up by a factor of ~40x for large problems and only a factor of 3-6x for smaller problems. The improved accuracy of double precision arithmetic was found to outweigh the speed improvement from faster calculations.

@@

A number of simple modifications to DOCTORS could be made that would greatly increase the code's usability and robustness. One of these modifications, generation of more refined group structures, is not a change to DOCTORS \textit{per se} but rather a change in the input cross section data. The other changes are additions to DOCTORS that would add new capabilities to the software. In addition to specific modifications, some additional, broader future goals can also be identified.

\subsection{Anisotropy treatment}
A more sophisticated anisotropy treatment would be appropriate since the collided flux was found to have issues regarding its behavior. Alternatively, adding a spherical harmonics solver may help the anisotropy treatment as well, or at least reduce its memory footprint and runtime. This would require moving the solution from discrete angle space to flux moment space which can greatly complicate the debugging. However, once the correctness of the overall algorithm is shown, this step should be relatively simple.

\subsection{Group structure}
The cross section data currently used by DOCTORS are taken from SCALE6.2. While these cross sections have been found to be sufficient to produce flux distributions in medical CT imaging, a more refined group structure designed for medical applications would be worth investigation. Also, the data distribtuion from SCALE6.2 contains potentially export controlled information since it also includes nuclear reactor materials. A medical purposed cross section library would alleviate this problem and allow the code to be freely released with a dataset. In the current version of DOCTORS, the user is required to obtian the cross section data files independently. This work can be done using either NJOY or the newly released AMPX code. Either code has the capability to collapse an ENDF formatted data file into a group structure of the user's choosing in the format DOCTORS reads.

\subsection{Therapy extension}
An advantage of the raytracer is that it is very fast. Therefore, it can characterize many beams and integrate them temporally easily. This would allow DOCTORS to model more complex beam shapes and scan protocols accurately. This would be particularly useful for medical treatment systems that use a multileaf collimator.

Some treatment systems employ a multileaf collimator to continuously shape the beam, resulting in a large dose deposition only at the area of interest. DOCTORS can likely be extended to high energy therapeutic beams for clinical treatment. However, higher energy photons will scatter more anisotropicly. This will require careful analysis of the angular treatment used.

\subsection{Partial acceptance criteria}
In the current version of doctors, the raytracer is very accurate with respect to MCNP6 in predicting the uncollided flux, except for along the periphery of the beam. Voxels are currently either completely inside the beam or completely outside of it as determined by its isocenter. A more sophisticated technique whereby a voxel cah be partially accepted in the beam would remove these artifacts.

\subsection{Organ identification}
Currently, DOCTORS cannot automatically identify specific organs, thus it is only able to compute the equivalent dose via dose deposition. If specific organs could be identified, tissue specific weighting factors could be applied resulting in the effective dose to each organ which would be of greater clinical significance. However, such identification currently requires additional input about the scan protocol and extensive knowledge about the human anatomy not currently integrated into DOCTORS.
